---
title: "day-26"
format: html
---


Load Libraries
```{r}
library(ggpubr)
library(raster)
library(tidyverse)
library(readr)
library(terra)
library(glue)
```


Read in Data. To enable terra::rast() in R to read a raster file directly from a GitHub repository using a VSI-prefixed URL
```{r}
url <- '/vsicurl/https://raw.githubusercontent.com/mikejohnson51/csu-ess-330/main/resources/foco-elev-cm.tif'

r <- terra::rast(url)

print(r)
```

Once the date is read in, describe the structure of the data in a few sentences:

The structure of the data is in a spatRaster class and it has a maximum value of 197781 and a minimum value of 146730. The dimensions are 725 rows, 572 columns, and 1 nlayer. The data is currently in a NAD83 datum that uses the units m. 


Natively, the data has a horizontal unit of cm. Multiply the data by 0.0328084 to convert the data to feet.
```{r}
#  x and y units are in meters, not the elevation values. convert elev from cm-->ft
r_ft <- r * 0.0328084
```


Extract the values from the raster into a data.frame by using the dataframe=TRUE parameter in the values() function.
```{r}
#gives you a data frame with each row representing a raster cell and its corresponding elevation in feet.

 #Extract raster values into a data frame (one column of elevation values)
df <- terra::values(r_ft, dataframe = TRUE)
head(df)
```

Create a density plot using ggpubr and save it with ggsave
```{r}
# Create a density plot using ggpubr
p <- ggdensity(df, x = names(df),      # Use the column of elevation values
               fill = "lightblue",     # Fill color for the density curve
               color = "black",        # Border color of the curve
               add = "mean",           # Add a vertical line for the mean
               rug = TRUE) +           # Add a rug plot (ticks) along the x-axis
  labs(title = "Elevation Density Plot",   # Plot title
       x = "Elevation (feet)",             # X-axis label
       y = "Density") +                    # Y-axis label
  theme_pubr()                             # Apply a clean ggpubr theme

# Save plot as PNG
ggsave("elevation_density.png", plot = p, width = 6, height = 4, dpi = 300)
```

